Topological Equivariant Artist Model

### 2025-04-07 paper/main.tex
spivak section rework

\paragraph{Joining indexing spaces: $\oplus: \mathcal{\dbasec} \sqcup \mathcal{\dbasec} \rightarrow \mathcal{\dbasec}$}
Joining two bundles on their base space, for example to glue the two trivializations in \autoref{fig:atct:fb} into a bundle, is the coproduct $\dbase_{a} \sqcup_{\dbase_{c}} \dbase_{b}$ over a space $\dbase_{c}$. A property of the coproduct is that the inclusion morphism must be commutative, which means that index spaces are combined on overlapping indices:
\begin{figure}[h]
  \includegraphics*[width=1\columnwidth]{figures/tex/k_coproduct.pdf}  \caption{Index spaces are combined via the \textit{coproduct} $\dbase_{a}\sqcup_{\dbase_c} \dbase_{b}$.}
  \label{fig:atct:base}
\end{figure}
 For example, in \autoref{fig:atct:base}, the open interval $\dbase_c$ is present in both $\dbase_a$ and $\dbase_c$; therefore $\dbase_a$ and $\dbase_b$ are glued together where they overlap at $\dbase_c$.  Two spaces have been combined correctly if every point in the overlap  $\dbasepoint \in \dbase^{c}$ is present in each of the spaces $\dbasepoint \in \dbase^{a}$ and $\dbasepoint \in \dbase^{b}$ such that it is present in the joined space $\dbasepoint \in \dbase^{a}\sqcup_{\dbase^{c}} \dbase_{b}$. This simple test that the records are joined correctly is what allows us to reliably build larger datasets out of smaller ones, such as in the case of distributed and on demand datasets.

@@

Spivak introduces a \textit{type specification} as a bundle map $\pi: \mathscr{U} \rightarrow \textbf{DT}$. The base space $\textbf{DT}$ is a set of data types $T \in \textbf{DT}$ and the total space $\mathscr{U}$ is the disjoint union of the domains of each type $\mathscr{U} = \bigsqcup_{T \in \textbf{DT}} \pi^{-1}(T)$ such that each element $x$ in the domain $\pi^{-1}(T)$ is one possible value of an object of type $T$ \cite{spivakSimplicialDatabases2009}. For example, if $T=\texttt{int}$, then the image $\pi^{-1}(\texttt{int}) = \mathbb{Z} \subset \mathscr{U}$ is the set of all integers and $x=3 \in \mathbb{Z}$ is the value of one $\texttt{int}$ object. Spivak's analogue of a database \textit{schema}\cite{ullmanFirstCourseDatabase2008} is the \textit{simple schema of type $\pi$} which is defined as consisting of the pair $(\fnames, \sigma)$ where $\fnames$ is the set of field names and $\sigma: \fnames \rightarrow \textbf{DT}$ is a function from field name to field data type\cite{spivakSimplicialDatabases2009}. The function $\sigma$ is composed with $\pi$ such that $\pi^{-1}(\sigma(C)) \subseteq \mathscr{U}$; this composition induces a domain bundle $\pi_{\sigma}:\mathscr{U}_{\sigma} \rightarrow \fnames$ that associates a field name $c \in C$ with its corresponding domain $\pi^{-1}_{\sigma}(C) \subseteq \mathscr{U}_{\sigma}$. He defines a \textit{record} as a function $\delement: \fnames \rightarrow \mathscr{U}_{\sigma}$ from column name to type domain and the set of records on $\pi_{\sigma}$ is denoted $\Gamma^{\pi}(\sigma)$. Records must return an object of type $\sigma(\fname) \in \textbf{DT}$ for each field $c \in C$. Spivak then describes tables as sections $\dsection: \dbase \rightarrow \Gamma^{\pi}(\sigma)$ from an indexing space $\dbase$ to the set of all possible records $\Gamma^{\pi}(\sigma)$ on the schema bundle, and his notion of a table generalizes to our notion of a data container.

In summary, Spivak defines types as maps between type name (e.g. $temperature \in C$), data type ($\texttt{float} \in \texttt{DT}$), and type domain $\mathbb{R}\subseteq \mathscr{U_{\sigma}} \subseteq \mathscr{U}$). Records are maps from name to value ($\textit{temperature}:152.0^{\circ} F$) and tables are maps from index to record \cite{spivakSimplicialDatabases2009}:
\begin{table}[H]
  \begin{equation*}
    \begin{tabular}{|c|c|c|}
      \hline
       types and schema & \textcolor{fiber}{record} & \textcolor{section}{table}\\
      \hline
      \begin{tikzcd}[ampersand replacement=\&, sep=huge]
        \textcolor{total}{\mathcal{U}_\sigma} \arrow[d, "\pi_{\sigma}"', color=total, two heads] \arrow[r, dotted] \arrow[rd, dotted] \& \textcolor{total}{\mathcal{U}} \arrow[d, "\pi"', color=total, two heads] \\
        \textcolor{base}{C} \arrow[r, "\sigma"']                                                                                   \& \textcolor{base}{\texttt{DT}}
      \end{tikzcd}
    &
    \begin{tikzcd}[ampersand replacement=\&, sep=huge]
      \textcolor{total}{\mathcal{U}_{\sigma}} \arrow[d, "\pi_{\sigma}"', color=total, two heads] \\
      \textcolor{base}{C} \arrow[u, "\delementc"', dashed, bend right, color=fiber]
      \end{tikzcd}
    &
    \begin{tikzcd}[ampersand replacement=\&, sep=huge]
      \dfiberc\coloneqq \textcolor{total}{\Gamma^{\pi}(\sigma)} \arrow[d, "\pi"', color=total, two heads] \\
      \dbasec \arrow[u, "\dsectionc"', dashed, bend right, color=section]
      \end{tikzcd}\\
      \hline
    \end{tabular}
    \end{equation*}
    \caption{}
    \label{tab:fiber_specification}
  \end{table}

Building on Spivak, we define the \textcolor{fiber}{fiber space} \dfiberc\ to be the space of all possible data records:
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$. Adopting Spivak's fiber bundle construction of types allows our model to reuse types so long as the field names are distinct and that field values can be accessed by field name,  since those are sections on $\mathcal{U}_{\sigma}$. Furthermore, since domains $\mathscr{U}_{{\sigma}}$ of types are a mathematical space, multi-dimensional fields can be encoded in the same manner as single dimensional fields and fields can have different names but the same type.

@@

\paragraph{Merging fields: $\otimes: \mathcal{\dfiber} \times \mathcal{\dfiber} \rightarrow \mathcal{\dfiber}$ }

@@

Combining fibers correctly can be verified by checking that when a fiber component $\dfiber^{c}$ is present in both $\dfiber^{a}$ and $\dfiber^{b}$, it is identical when projected out of either such that the \textit{product} diagram commutes, means that when data with many fields is decomposed into its component fields, records maintain their integrity:

\begin{figure}[H]
  \includegraphics*[width=1\columnwidth]{figures/tex/f_product.pdf}
  \caption{Fields are combined via the \textit{product} $\dfiber_{a}\times_{\dfiber_c} \dfiber_{b}$.}
  \label{fig:atct:fiber}
\end{figure}

For example in \autoref{fig:atct:fiber} the red (temperature, time, pressure) record separates into (temperature, time) and (pressure, time) records that share the same red time. Furthermore this time is the same whether it is obtained from the (temperature, time) or (pressure, time) record. Generally, the test for joining fields is that when a record is present in a fiber $\delement_{c} \in \pi_c(\dfiber_{a}), \delement_c \in \pi_c(\dfiber_{b})$ then it is in the joint record $\delement_{c} \in \pi_c(\dfiber_{a}\times_{\dfiber_c}\dfiber_{b})$ when is in the field $\dfiber_c$ being joined on. This simple test that fields are joined together correctly for the same record is what allows us to reliably combine multiple datasets together on shared properties, for example growing the weather station data in \autoref{fig:related-work:continuity:ktypes} from a temporal to spatial dataset by adding the weather at each location at each time.

@@

\subsection{Composing Artists}\label{sec:artist:operators}
\note{addition: intersections mapped same, multiplication: fibers mapped same}
\note{large big data glued together correctly}
A common use of category theory in software engineering is the specification of modular components \cite{wielsManagementEvolvingSpecifications1998} such that we can build systems where the structure preserved by components is preserved in the composition of the components. This allows us to express that an artist that works on a dataset can be composed of artists that work on sub parts of that dataset.

>>>>

@@

Spivak introduces a \textit{type specification} as a bundle map $\pi: \mathscr{U} \rightarrow \textbf{DT}$. The base space $\textbf{DT}$ is a set of data types $T \in \textbf{DT}$ and the total space $\mathscr{U}$ is the disjoint union of the domains of each type $\mathscr{U} = \bigsqcup_{T \in \textbf{DT}} \pi^{-1}(T)$ such that each element $x$ in the domain $\pi^{-1}(T)$ is one possible value of an object of type $T$ \cite{spivakSimplicialDatabases2009}. For example, if $T=\texttt{int}$, then the image $\pi^{-1}(\texttt{int}) = \mathbb{Z} \subset \mathscr{U}$ is the set of all integers and $x=3 \in \mathbb{Z}$ is the value of one $\texttt{int}$ object. Spivak's analogue of a database \textit{schema}\cite{ullmanFirstCourseDatabase2008} is the \textit{simple schema of type $\pi$} which is defined as consisting of the pair $(\fnames, \sigma)$ where $\fnames$ is the set of field names and $\sigma: \fnames \rightarrow \textbf{DT}$ is a function from field name to field data type\cite{spivakSimplicialDatabases2009}.

The function $\sigma$ is composed with $\pi$ such that $\pi^{-1}(\sigma(C)) \subseteq \mathscr{U}$; this composition induces a domain bundle $\pi_{\sigma}:\mathscr{U}_{\sigma} \rightarrow \fnames$ that associates a field name $c \in C$ with its corresponding domain $\pi^{-1}_{\sigma}(C) \subseteq \mathscr{U}_{\sigma}$. He defines a \textit{record} as a function $\delement: \fnames \rightarrow \mathscr{U}_{\sigma}$ from column name to type domain and the set of records on $\pi_{\sigma}$ is denoted $\Gamma^{\pi}(\sigma)$. Records must return an object of type $\sigma(\fname) \in \textbf{DT}$ for each field $c \in C$. Spivak then describes tables as sections $\dsection: \dbase \rightarrow \Gamma^{\pi}(\sigma)$ from an indexing space $\dbase$ to the set of all possible records $\Gamma^{\pi}(\sigma)$ on the schema bundle, and his construction of a table generalizes to our notion of a dataset.

In summary, Spivak defines \textit{types} as map $\sigma$ from type name (e.g. $temperature \in C$) to data type ($\texttt{float} \in \texttt{DT}$) and maps $\pi_{\sigma}, \pi$ from type domain ($\mathbb{R}\subseteq \mathscr{U_{\sigma}} \subseteq \mathscr{U}$) to name and type, \textit{schemas} as maps from type name to data type ($temperature: \texttt{float}$), \textit{records} as maps from name to value ($temperature:152.0^{\circ} F$) and \textit{tables} as maps from index to record \cite{spivakSimplicialDatabases2009}:
\begin{equation*}
    \begin{tabular}{|c|c|c|}
      \hline
       types and schema & \textcolor{fiber}{record} & \textcolor{section}{table}\\
      \hline
      \begin{tikzcd}[ampersand replacement=\&, sep=huge]
        \textcolor{total}{\mathcal{U}_\sigma} \arrow[d, "\pi_{\sigma}"', color=total, two heads] \arrow[r, dotted] \arrow[rd, dotted] \& \textcolor{total}{\mathcal{U}} \arrow[d, "\pi"', color=total, two heads] \\
        \textcolor{base}{C} \arrow[r, "\sigma"']                                                                                   \& \textcolor{base}{\texttt{DT}}
      \end{tikzcd}
    &
    \begin{tikzcd}[ampersand replacement=\&, sep=huge]
      \textcolor{total}{\mathcal{U}_{\sigma}} \arrow[d, "\pi_{\sigma}"', color=total, two heads] \\
      \textcolor{base}{C} \arrow[u, "\delement"', dashed, bend right, color=fiber]
      \end{tikzcd}
    &
    \begin{tikzcd}[ampersand replacement=\&, sep=huge]
      \dfiberc\coloneqq \textcolor{total}{\Gamma^{\pi}(\sigma)} \arrow[d, "\pi"', color=total, two heads] \\
      \dbasec \arrow[u, "\dsectionc"', dashed, bend right, color=section]
      \end{tikzcd}\\
      \hline
    \end{tabular}
\end{equation*}

Building on Spivak's constructions, we define the \textcolor{fiber}{fiber space} \dfiberc\ to be the space of all possible data records:
\begin{equation}
  \dfiberc \coloneqq \{\delement: \fnames \rightarrow \mathscr{U}_{\sigma} \bigm{\vert} \pi_{\sigma}(\delement(\fname)) = \fname\;for\;all\; \fname \in \fnames \}
\end{equation}
such that the preimage of a point is the corresponding data type domain $\pi^{-1}(\dbasepoint) = \dfiber_{k} = \mathscr{U}_{{\sigma}_{\dbasepoint}}$. Adopting Spivak's fiber bundle construction of types allows our model to reuse types so long as the field names are distinct and that field values can be accessed by field name,  since those are sections on $\mathcal{U}_{\sigma}$. Furthermore, since domains $\mathscr{U}_{{\sigma}}$ of types are a mathematical space, multi-dimensional fields can be encoded in the same manner as single dimensional fields and fields can have different names but the same type.

@@

@@

\subsubsection{Adding and expanding records}

A common use of category theory in software engineering is the specification of modular components \cite{wielsManagementEvolvingSpecifications1998} such that we can build systems where the structure preserved by components is preserved in the composition of the components. As described in \autoref{sec:atct:fb:base}, the index space is modeled as a set of open sets; therefore adding index spaces $\dbase_a$ and $\dbase_b$ is the coproduct $\dbase_{a} \sqcup_{\dbase_{c}} \dbase_{b}$ over a shared space $\dbase_{c}$. As described in \autoref{sec:atct:fb:fiber}, the bifunctor can be used to build composite fiber types;  in the same manner fields $\dfiber_a$ and $\dfiber_b$ can be composed by taking the fiber product $\dfiber_{a}\times_{\dfiber_{c}} \dfiber_b$ over a shared fiber $\dfiber_c$.

\begin{figure}[H]
  \includegraphics*[width=1\columnwidth]{figures/tex/more_records.pdf}
  \caption{Records are added to the dataset as a \textit{coproduct} $\dbase_{a}\sqcup_{\dbase_c} \dbase_{b}$ of index spaces. Fields are added to each record as a \textit{product} $\dfiber_{a}\times_{\dfiber_c} \dfiber_{b}$ of fiber spaces.}
  \label{fig:atct:joint}
\end{figure}

For example, in \autoref{fig:atct:joint}, the open interval $\dbase_c$ is present in both $\dbase_a$ and $\dbase_c$; therefore $\dbase_a$ and $\dbase_b$ are glued together where they overlap at $\dbase_c$.  Two spaces have been combined correctly if every point in the overlap  $\dbasepoint \in \dbase^{c}$ is present in each of the spaces $\dbasepoint \in \dbase^{a}$ and $\dbasepoint \in \dbase^{b}$ such that it is present in the joined space $\dbasepoint \in \dbase^{a}\sqcup_{\dbase^{c}} \dbase_{b}$. In \autoref{fig:atct:joint}, the red (temperature, time, pressure) record separates into (temperature, time) and (pressure, time) records that share the same red time. Furthermore this time is the same whether it is obtained from the (temperature, time) or (pressure, time) record. Generally, the test for joining fields is that when a record is present in a fiber $\delement_{c} \in \pi_c(\dfiber_{a}), \delement_c \in \pi_c(\dfiber_{b})$ then it is in the joint record $\delement_{c} \in \pi_c(\dfiber_{a}\times_{\dfiber_c}\dfiber_{b})$ when is in the field $\dfiber_c$ being joined on. Verifying that indicies are joined correctly allows us to reliably build larger datasets out of smaller ones, such as in the case of distributed and on demand datasets. Verifying that fields are joined together correctly for the same record allows us to reliably combine multiple datasets together on shared properties, for example growing the weather station data in \autoref{fig:related-work:continuity:ktypes} from a temporal to spatial dataset by adding the weather at each location at each time.

@@

\subsection{Composing Artists}\label{sec:artist:operators}
\note{addition: intersections mapped same, multiplication: fibers mapped same}
\note{large big data glued together correctly}
 This allows us to express that an artist that works on a dataset can be composed of artists that work on sub parts of that dataset.
